---
title: "Test election tracker"
author: "John Ray"
date: "November 7, 2017"

layout: post

output: 
  md_document:
    variant: markdown_github
---

On this page, I am tracking election results for the State House, State Senate/State Assembly, Governor, and Congressional races going on today.

```{r include=FALSE, warn=FALSE, echo=FALSE}
graphics.off()
rm(list = ls())

library(dplyr)
library(ggplot2)
library(knitr)
library(plotly)
library(ggiraph)
library(gridExtra)

load('~/Dropbox/consulting/election_tracker/nightof_data/historical_plots.RData')

gov <- read.csv('~/Dropbox/consulting/election_tracker/nightof_data/gov.csv', stringsAsFactors = F)
#cd <- read.csv('~/Dropbox/consulting/election_tracker/nightof_data/cd.csv', stringsAsFactors = F)
sh <- read.csv('~/Dropbox/consulting/election_tracker/nightof_data/sh.csv', stringsAsFactors = F)
ss <- read.csv('~/Dropbox/consulting/election_tracker/nightof_data/ss.csv', stringsAsFactors = F)

gov$new_margin = gov$dem2017/(gov$dem2017 + gov$rep2017 + .001)
#cd$new_margin = cd$dem2017/(cd$dem2017 + cd$rep2017 + 1)
sh$new_margin = sh$dem2017/(sh$dem2017 + sh$rep2017 + .001)
ss$new_margin = ss$Dem2017/(ss$Dem2017 + ss$Rep2017 + .001)

#cd_lines = dplyr::left_join(cd_lines, cd, by = c('STATEFP', 'CD115FP'))
county_df = dplyr::left_join(county_df, gov, by = c('STATEFP','NAME'))
```

```{r echo=FALSE, warn=FALSE}
#gov
gov_gg_new <- ggplot(county_df, aes(x = long, y = lat, group = group, fill = new_margin)) +
  geom_polygon(color = 'black') + 
  theme_void() +
  scale_fill_continuous(low = 'red', high = 'darkblue', name = c("Dem %"), labels = scales::percent, limits = c(0, 1), breaks = c(0, 1)) +
  theme(panel.background = element_rect(color = 'black' , size = 1), legend.position = 'bottom') +
  labs(title = 'Governor, 11/7/17')

gov_bar_old <- ggplot(data = gov, aes(x = NAME)) +
  geom_point(aes(y = prev_margin, color = prev_margin), position = 'identity') +
  geom_segment(aes(xend = NAME, y = 0, yend = prev_margin, color = prev_margin)) +
  theme_classic() +
  scale_color_continuous(low = 'red', high = 'darkblue') +
  labs(y = '', x = '', title = 'Margin by district/county, previous election') +
  scale_y_continuous(labels = scales::percent) +
  theme(panel.background = element_rect(color = 'black', fill = NA, size = 1), legend.position = 'null', axis.text.x = element_text(angle = 90, hjust = 1, size = 4)) +
  facet_wrap(~state)

gov_bar_new <- ggplot(data = gov, aes(x = NAME)) +
  geom_point(aes(y = new_margin, color = new_margin), position = 'identity') +
  geom_segment(aes(xend = NAME, y = 0, yend = new_margin, color = new_margin)) +
  theme_classic() +
  scale_color_continuous(low = 'red', high = 'darkblue', breaks = c(-.5, .5)) +
  labs(y = '', x = '', title = 'Margin by district/county, 11/7/17') +
  scale_y_continuous(labels = scales::percent, breaks = c(-1, 0, 1)) +
  theme(panel.background = element_rect(color = 'black', fill = NA, size = 1), legend.position = 'null', axis.text.x = element_text(angle = 90, hjust = 1, size = 4)) +
  facet_wrap(~state)

#ss
ss_gg_new <- ggplot(data = ss, aes(xmin = x - .5, xmax = x + .5, ymin = y - .25, ymax = y + .25, fill = new_margin, label = District)) +
  geom_rect_interactive() + 
  theme_void() +
  scale_fill_continuous(low = 'red', high = 'darkblue') +
  labs(title = 'State Senate') +
  theme(panel.background = element_rect(color = 'black', fill = NA, size = 1), legend.position = 'null')

ss_bar_old <- ggplot(data = ss, aes(x = NAME)) +
  geom_point(aes(y = prev_margin, color = prev_margin), position = 'identity') +
  geom_segment(aes(xend = NAME, y = 0, yend = prev_margin, color = prev_margin)) +
  theme_classic() +
  scale_color_continuous(low = 'red', high = 'darkblue') +
  labs(y = '', x = '') +
  scale_y_continuous(labels = scales::percent) +
  theme(panel.background = element_rect(color = 'black', fill = NA, size = 1), legend.position = 'null', axis.text.x = element_text(angle = 90, hjust = 1, size = 4)) +
  facet_wrap(~state)

ss_bar_new <- ggplot(data = ss, aes(x = NAME)) +
  geom_point(aes(y = new_margin, color = new_margin), position = 'identity') +
  geom_segment(aes(xend = NAME, y = 0, yend = new_margin, color = new_margin)) +
  theme_classic() +
  scale_color_continuous(low = 'red', high = 'darkblue') +
  labs(y = '', x = '') +
  scale_y_continuous(labels = scales::percent, breaks = c(-.5, 0, .5)) +
  theme(panel.background = element_rect(color = 'black', fill = NA, size = 1), legend.position = 'null', axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_wrap(~state)

#sh
sh_gg_new <- ggplot(data = sh, aes(xmin = x - .25, xmax = x + .25, ymin = y - .125, ymax = y + .125, fill = new_margin, label = District)) +
  geom_rect() + 
  theme_void() +
  scale_fill_continuous(low = 'red', high = 'darkblue') +
  labs(title = 'State House') +
  theme(panel.background = element_rect(color = 'black', fill = NA, size = 1), legend.position = 'null')

sh_bar_old <- ggplot(data = sh, aes(x = NAME)) +
  geom_point(aes(y = prev_margin, color = prev_margin), position = 'identity') +
  geom_segment(aes(xend = NAME, y = 0, yend = prev_margin, color = prev_margin)) +
  theme_classic() +
  scale_color_continuous(low = 'red', high = 'darkblue') +
  labs(y = '', x = '') +
  scale_y_continuous(labels = scales::percent, breaks = c(-1, 0, 1)) +
  theme(panel.background = element_rect(color = 'black', fill = NA, size = 1), legend.position = 'null', axis.text.x = element_text(angle = 90, hjust = 1, size = 4)) +
  facet_wrap(~state)

sh_bar_new <- ggplot(data = sh, aes(x = NAME)) +
  geom_point(aes(y = new_margin, color = new_margin), position = 'identity') +
    geom_segment(aes(xend = NAME, y = 0, yend = new_margin, color = new_margin)) +
  theme_classic() +
  scale_color_continuous(low = 'red', high = 'darkblue') +
  labs(y = '', x = '') +
  scale_y_continuous(labels = scales::percent, breaks = c(-1, 0, 1)) +
  theme(panel.background = element_rect(color = 'black', fill = NA, size = 1), legend.position = 'null', axis.text.x = element_text(angle = 90, hjust = 1, size = 4)) +
  facet_wrap(~state)
```

### Tonight's results

The races I will be watching tonight include

* Virginia governorship and State House
* New Jersey governorship and General Assembly
* Georgia State House 4, 26, 42, 56, 60, 89, 117, and 119
* Georgia State Senate 6 and 39
* Maine State House 56
* Massachusetts 1st Berkshire and 3rd Essex (State House)
* Michigan State House 1 and 109
* Missouri State House 23 and 151
* Missouri State Senate 8
* New Hampshire Hillsborough 15 (State House)
* New York State Assembly 27 and 71
* New York State Senate 26
* South Carolina State House 113
* Washington State House 7, 31, and 48
* Washington State Senate 7, 37, 45, and 48

The first column of plots include the Governors' races. The second column of plots includes the lower house races, with lower house seats across the US drawn as equally-sized pixels. The third column includes upper house races.

I will be tracking, but not plotting, the Utah 3rd Congressional district race as well. I could not devise an optimal visual display that included UT-3, but I will note the results.

```{r echo = FALSE, warning = FALSE}
g1 = grid.arrange(gov_gg_new, sh_gg_new, ss_gg_new, ncol = 3)
ggsave(plot = g1, "~/johnlray.github.io/2017/11/07/figure/unnamed-chunk-3-1.png", height = 5.2, width = 16)
```

### Results from these districts' previous elections

```{r echo = FALSE, warning = FALSE}
g2 = grid.arrange(gov_gg, sh_gg, ss_gg, ncol = 3)
ggsave(plot = g2, "~/johnlray.github.io/2017/11/07/figure/unnamed-chunk-4-1.png", height = 5.2, width = 16)
```

### Tonight's Dem - Rep margins

```{r echo = FALSE, warning = FALSE}
g3 = grid.arrange(gov_bar_new, sh_bar_new, ss_bar_new, ncol = 3)
ggsave(plot = g3, "~/johnlray.github.io/2017/11/07/figure/unnamed-chunk-5-1.png", height = 5.2, width = 16)
```

### These districts' previous Dem - Rep margins

```{r echo = FALSE, warning = FALSE}
g4 = grid.arrange(gov_bar_old, sh_bar_old, ss_bar_old, ncol = 3)
ggsave(plot = g4, "~/johnlray.github.io/2017/11/07/figure/unnamed-chunk-6-1.png", height = 5.2, width = 16)
```

Here is the data that goes into these graphs.

```{r echo = FALSE, warning = FALSE}
gov_print <- gov[,c('state','NAME','prev_margin','new_margin')]
colnames(gov_print) <- c('State','County','Previous margin','Tonights margin')

sh_print <- sh[,c('state','NAME','prev_margin','new_margin')]
colnames(sh_print) <- c('State','County','Previous margin','Tonights margin')

ss_print <- ss[,c('state','NAME','prev_margin','new_margin')]
colnames(ss_print) <- c('State','County','Previous margin','Tonights margin')

kable(gov_print, caption = "Governors race data")
kable(ss_print, caption = "Upper chamber race data")
kable(sh_print, caption = "Lower chamber race data")
```