---
title: "What is the Z-Transformation Doing?"
author: "John Ray"
date: "February 6, 2017"
output: html_document
---

### What is the Z-Transformation Doing?

In the statistics class I teach here at UCLA, I find students are often confused by one of the most important parts of statistics: the Z transformation. From a pedagogical standpoint, this is understandable. I am generally nervous every time I try to teach the Z transformation. Not only did I learn it in a confusing way on my first few passes through statistics, but I have tried several different ways of constructing the concept for my students in a way that will most easily help them encode it into their memories, and I don't feel I've quite cracked it yet.

Here, I try a different approach to presenting the purpose of the Z-transformation: I start at the end and work my way backward. I do this because the ``end'' of the process is the part we can actually understand the most easily, and the previous stuff is almos purely conceptual.

For example, lets start with a question we already know the answer to: Do Democrats and Republicans feel differently about President Trump? The answer, of course, is yes. Not just yes. *OBVIOUSLY, YES!* Before even asking students to sit down and work through some math, I'm presenting them with a problem they already know the answer to.

What is ``the problem'' here? Its so straightforward it almost seems possible to gloss over it. But lets be specific: I've said that Democrats and Republicans feel *differently* about President Trump. What's specific about that? Well, if we want, we can think of that question as a hypothesis.

If Democrats and Republicans feel differently about Trump, then if I go and sample some Democrats and sample some Republicans, I will find that the *average* approval of Trump by Democrats and the *average* approval of Trump by Republicans will be different. Who cares about the average? Well, trying to draw inferences from the opinions of individuals is cumbersome, unless I try and summarize the opinions of large numbers of people. For that purpose, I take the *mean* of my samples.

How good are my summary statistics? Well, the means of my samples are only as good as those means are *precise.* That is, if I sample Democrats, if they all feel about the same about Trump as each other, then the mean will be a very good summary statistic. But if I sample Democrats and it turns out that as many Democrats love Trump as hate Trump, then the mean is not a very meaningful number, because the *spread* around the mean will be huge. We measure the value of the *spread* using another statistic that we call the *standard deviation.*

So lets say I go out and ask a bunch of Democrats how they feel about Trump, and a bunch of Republicans about Trump. Using the latest survey data, I would find that the Democrats look something like

```{r}
dems = c(rep(0, 140), rep(1, 60))
```

and the Republicans look something like

```{r}
reps = c(rep(0, 20), rep(1, 180))
```

When I say "what the data look like," I as a data scientist am mostly interested in just a few things:

* Some properties of the distribution of each variable (mean, median, mode, min, max)
* The overlap between the variables
* Some measure of how the variables differ